---
title: "432 Class 23"
author: Thomas E. Love, Ph.D.
date: "2025-04-08"
format:
  revealjs: 
    theme: dark
    embed-resources: true
    self-contained: true
    slide-number: true
    footnotes-hover: true
    preview-links: auto
    date-format: iso
    logo: 432-2025-pic.png
    footer: "432 Class 23 | 2025-04-08 | <https://thomaselove.github.io/432-2025/>"
---

## Today's Topic

**Cox models for time-to-event data**

- Rossi's study of recidivism in people released from Maryland state prisons
- Reviewing what we've done so far using `coxph()`
- Using `cph()` from `rms` to fit three different Cox models

## Setup

```{r}
#| echo: true

knitr::opts_chunk$set(comment=NA)

library(janitor)
library(naniar)
library(here)
library(conflicted)
library(broom)
library(gt)
library(rms)
library(survival)
library(survminer)
library(easystats)
library(tidyverse)

theme_set(theme_bw())
```

## The Rossi Data

This data set is originally from Rossi et al. (1980). The data pertain to 432 convicts who were released from Maryland state prisons in the 1970s and who were followed up for one year after release. Half the released convicts were assigned at random to an experimental treatment in which they were given financial aid; half did not receive aid. Details on Rossi data variables and descriptions are provided [at this link](https://rdrr.io/cran/carData/man/Rossi.html). 

- We will use a subset of the available variables today. 
- Subject ID code (`s_id`) included in the `rossi.csv` file.

## Variable Descriptions for `rossi_432` {.smaller}

Variable | Description
:-------: | :------------------------------------------------------
`s_id` | Subject ID code
`week` | week of first arrest after release or censoring; all censored observations are censored at 52 weeks.
`arrest` | 1 if arrested, 0 if not
`fin` | financial aid: no or yes (main treatment of study)
`age` | age in years at time of release
`race` | two categories: black or other
`wexp` | full-time work experience before incarceration: no or yes
`prio` | number of convictions prior to current incarceration

## Ingesting the data into `rossi_432`

The `rossi.csv` file on our 432 data page includes 432 observations on 63 variables, but we'll look at just the 8 we listed on the last slide. 

```{r}
#| echo: true

rossi_432 <- read_csv(here("c23/data/rossi.csv"), show_col_types = FALSE) |>
  select(s_id, week, arrest, fin, age, race, wexp, prio)  |>
  mutate(across(where(is.character), as_factor),
         s_id = as.character(s_id))

dim(rossi_432)
n_miss(rossi_432)
```

## The `rossi_432` tibble

```{r}
#| echo: true
rossi_432
```

## Create a survival object

```{r}
#| echo: true
 
rossi_432$S <- with(rossi_432, Surv(week, arrest == 1))

head(rossi_432$S)
```

All subjects were followed either for 52 weeks after their release, or until they were (re-)arrested. So all "censored" subjects will have 52 weeks without re-arrest.

### Compare the two financial aid groups

```{r}
#| echo: true

km_fin <- survfit(S ~ fin, data = rossi_432)

km_fin
```

## Plot the K-M curves

```{r}
#| echo: true
ggsurvplot(km_fin, data = rossi_432, 
           palette = "aaas", ggtheme = theme_bw(),
           pval = TRUE)
```

## Log - Log Plot for K-M estimation

- Do these curves cross?

```{r}
#| echo: true
#| fig-height: 4
plot(survfit(S ~ fin, data = rossi_432), col = c(1:2), fun = "cloglog")
```

## Cumulative Hazard Plot

```{r}
#| echo: true
#| output-location: slide

ggsurvplot(km_fin, data = rossi_432, fun = "cumhaz",
           xlab = "Weeks of Follow-Up", palette = "aaas",
           pval = TRUE, break.time.by = 13,
           risk.table = TRUE, risk.table.height = 0.25)
```

## Cox Model using `fin` only

```{r}
#| echo: true

fit1 <- coxph(S ~ fin, data = rossi_432)

fit1
```

- How do we interpret the hazard ratio estimate of 0.6914?

## `fit1` model parameters

```{r}
#| echo: true

model_parameters(fit1, pretty_names = FALSE, exponentiate = TRUE,
                 ci = 0.90, digits = 3)
```

## Model `fit1` summaries

```{r}
#| echo: true

model_performance(fit1)
```

```{r}
#| echo: true

glance(fit1) |> gt() |> fmt_number(decimals = 3)
```

## Forest Plot for `fit1`

```{r}
#| echo: true

ggforest(fit1)
```

## Checking PH Assumption

```{r}
#| echo: true

cox.zph(fit1)
```

## Checking PH Assumption

```{r}
#| echo: true

ggcoxzph(cox.zph(fit1))
```

## A larger Cox model (`fit2`)

```{r}
#| echo: true

fit2 <- coxph(S ~ fin + age + race + wexp + prio, data = rossi_432)

fit2
```

## `fit2` model parameters

```{r}
#| echo: true

model_parameters(fit2, pretty_names = FALSE, exponentiate = TRUE,
                 ci = 0.90, digits = 3)
```

- How do we interpret the `fin` hazard ratio estimate of 0.694?
- How about the `prio` hazard ratio estimate of 1.096?

## Model `fit2` summaries

```{r}
#| echo: true

model_performance(fit2)
```

```{r}
#| echo: true

glance(fit2) |> gt() |> fmt_number(decimals = 3)
```


## Checking PH Assumption

```{r}
#| echo: true

cox.zph(fit2)
```

## Checking PH Assumption

```{r}
#| echo: true

ggcoxzph(cox.zph(fit2))
```

## Compare our two models

```{r}
#| echo: true

plot(compare_performance(fit1, fit2))
```

## Using `cph()` from the **rms** package

```{r}
#| echo: true

dd <- datadist(rossi_432); options(datadist = "dd")

fit1_cph <- cph(S ~ fin, data = rossi_432,
                x = TRUE, y = TRUE, surv = TRUE)

fit2_cph <- cph(S ~ fin + age + race + wexp + prio, data = rossi_432,
                x = TRUE, y = TRUE, surv = TRUE)
```

## `fit1_cph` results

```{r}
#| echo: true

fit1_cph
```

## `fit1_cph` Effects Plot

```{r}
#| echo: true
#| fig-height: 3

plot(summary(fit1_cph))
```

```{r}
#| echo: true

summary(fit1_cph)
```

## Proportional Hazards Assumption?

```{r}
#| echo: true

cox.zph(fit1_cph, transform = "km", global = TRUE)
```

## Proportional Hazards Assumption?

```{r}
#| echo: true

ggcoxzph(cox.zph(fit1_cph, transform = "km", global = TRUE))
```

## Bootstrap Validated `fit1_cph` summaries

```{r}
#| echo: true

set.seed(432)
validate(fit1_cph, B = 300)
```

- Validated C statistic = 0.5 + (0.0898/2) = 0.5449
- Validated Nagelkerke $R^2$ = 0.0077

## `survplot()` from **rms** `fit1_cph`

```{r}
#| echo: true
#| fig-height: 4

survplot(fit1_cph, fin = c("yes", "no"),
         time.inc = 13, type = "kaplan-meier",
         xlab = "Study Time in Weeks")
```

## `fit1_cph` Prediction Plot (1/3)

```{r}
#| echo: true
ggplot(Predict(fit1_cph))
```

## `fit1_cph` Prediction Plot (2/3)

```{r}
#| echo: true
ggplot(Predict(fit1_cph, fun = exp))
```

## `fit1_cph` Prediction Plot (3/3)

```{r}
#| echo: true
ggplot(Predict(fit1_cph, fun = plogis))
```

## `fit1_cph` Nomogram

```{r}
#| echo: true
#| output-location: slide
 
sv <- Survival(fit1_cph)
surv52 <- function(x) sv(52, lp = x)

plot(nomogram(fit1_cph, fun = surv52, funlabel = c("52 week survival")))
```

## `fit2_cph` results

```{r}
#| echo: true

fit2_cph
```

## `fit2_cph` Effects Plot

```{r}
#| echo: true

plot(summary(fit2_cph))
```

## `fit2_cph` Effects Table

```{r}
#| echo: true

summary(fit2_cph)
```

## `survplot()` from **rms** `fit2_cph`

- Looking at the `fin` effect

```{r}
#| echo: true
#| fig-height: 4

survplot(fit2_cph, fin = c("no", "yes"),
         time.inc = 13, type = "kaplan-meier",
         xlab = "Study Time in Weeks")
```

## `survplot()` from **rms** `fit2_cph`

- Looking at the `race` effect

```{r}
#| echo: true
#| fig-height: 4

survplot(fit2_cph, race = c("black", "other"),
         time.inc = 13, type = "kaplan-meier",
         xlab = "Study Time in Weeks")
```

## `survplot()` from **rms** `fit2_cph`

- Looking at the `age` effect

```{r}
#| echo: true
#| fig-height: 4

survplot(fit2_cph, age = c(15, 25, 35, 45),
         time.inc = 13, type = "kaplan-meier",
         xlab = "Study Time in Weeks")
```

## `survplot()` from **rms** `fit2_cph`

- Looking at the `wexp` effect

```{r}
#| echo: true
#| fig-height: 4

survplot(fit2_cph, wexp = c("no", "yes"),
         time.inc = 13, type = "kaplan-meier",
         xlab = "Study Time in Weeks")
```


## `survplot()` from **rms** `fit2_cph`

- Looking at the `prio` effect

```{r}
#| echo: true
#| fig-height: 4

survplot(fit2_cph, prio = c(0, 2, 8, 16),
         time.inc = 13, type = "kaplan-meier",
         xlab = "Study Time in Weeks")
```

## `fit2_cph` Prediction Plots (1/3)

```{r}
#| echo: true
ggplot(Predict(fit2_cph))
```

## `fit2_cph` Prediction Plots (2/3)

```{r}
#| echo: true
ggplot(Predict(fit2_cph, fun = exp))
```

## `fit2_cph` Prediction Plots (3/3)

```{r}
#| echo: true
ggplot(Predict(fit2_cph, fun = plogis))
```

## Proportional Hazards Assumption?

```{r}
#| echo: true

cox.zph(fit2_cph, transform = "km", global = TRUE)
```

## Proportional Hazards Assumption?

```{r}
#| echo: true

ggcoxzph(cox.zph(fit2_cph, transform = "km", global = TRUE))
```

## Bootstrap Validated `fit2_cph` summaries

```{r}
#| echo: true

set.seed(432)
validate(fit2_cph, B = 300)
```

- Validated C statistic = 0.5 + (0.2429/2) = 0.62145
- Validated Nagelkerke $R^2$ = 0.0538

## Compare Performance

```{r}
#| echo: true
plot(compare_performance(fit1_cph, fit2_cph))
```

## `fit2_cph` Nomogram

```{r}
#| echo: true
#| output-location: slide
 
sv <- Survival(fit2_cph)
surv52 <- function(x) sv(52, lp = x)

plot(nomogram(fit2_cph, fun = surv52, funlabel = c("52 week survival")))
```

## Add Non-Linear Terms?

- Notice use of `week` rather than `S` here...

```{r}
#| echo: true

plot(spearman2(week ~ fin + age + race + wexp + prio, data = rossi_432))
```

## Model `fit3_cph`

- We'll include a four-knot restricted cubic spline in `age`, and the interaction between `prio` and `wexp`.

```{r}
#| echo: true

fit3_cph <- cph(S ~ rcs(age, 4) + prio * wexp + fin + race, 
                data = rossi_432, x = T, y = T, surv = T)
```

## `fit3_cph`

```{r}
#| echo: true

fit3_cph
```

## `fit3_cph` Effects Plot

```{r}
#| echo: true

plot(summary(fit3_cph))
```

## `fit3_cph` Effects Table

```{r}
#| echo: true

summary(fit3_cph)
```

## `survplot()` from **rms** `fit3_cph`

- Looking at the `fin` effect

```{r}
#| echo: true
#| fig-height: 4

survplot(fit3_cph, fin = c("no", "yes"),
         time.inc = 13, type = "kaplan-meier",
         xlab = "Study Time in Weeks")
```

## `survplot()` from **rms** `fit3_cph`

- Looking at the `age` effect

```{r}
#| echo: true
#| fig-height: 4

survplot(fit3_cph, age = c(15, 25, 35, 45),
         time.inc = 13, type = "kaplan-meier",
         xlab = "Study Time in Weeks")
```


## `fit3_cph` Prediction Plots (raw)

```{r}
#| echo: true
ggplot(Predict(fit3_cph))
```

## `fit3_cph` Prediction Plots (plogis)

```{r}
#| echo: true
ggplot(Predict(fit3_cph, fun = plogis))
```

## Proportional Hazards Assumption?

```{r}
#| echo: true

ggcoxzph(cox.zph(fit3_cph, transform = "km", global = TRUE))
```

## Bootstrap Validated `fit3_cph` summaries

```{r}
#| echo: true

set.seed(432)
validate(fit3_cph, B = 300)
```

- Validated C statistic = 0.5 + (0.2417/2) = 0.62085
- Validated Nagelkerke $R^2$ = 0.0510

## Compare Performance

```{r}
#| echo: true
plot(compare_performance(fit1_cph, fit2_cph, fit3_cph))
```

## `fit3_cph` Nomogram

```{r}
#| echo: true
#| output-location: slide
 
sv <- Survival(fit3_cph)
surv52 <- function(x) sv(52, lp = x)

plot(nomogram(fit3_cph, fun = surv52, funlabel = c("52 week survival")))
```
